---
title: "Control Structures and Flow"
short-title: "Control Structures and Flow"
author: "James Balamuta"
short-author: "J Balamuta"
date: '`r format(Sys.Date(), "%B %d, %Y")`'      # Month DD, YYYY (Main Slide)
short-date: '`r format(Sys.Date(), "%m/%d/%Y")`' # MM/DD/YYYY (Lower Right)
institute: "University of Illinois at Urbana-Champaign"
short-institute: "UIUC"
department: "Department of Informatics, Statistics"           # Institute must be defined
license: "CC BY-NC-SA 4.0, 2016 - 2017, James J Balamuta"
section-titles: true                             # Provides slide headings
safe-columns: true                               # Enables special latex macros for columns.
output: 
   uiucthemes::beamer_illinois
---

### Announcements

- Groups!
- **Reminder** HW1 Due June 25th @ 11:59 PM CDT
- HW2 assigned tomorrow and due July 2nd @ 11:59 PM CDT
- Lecture Feedback Survey for Week #2 Available Now at: <https://goo.gl/forms/SQQlO3zZ9lc3j1o83>

### On the Agenda   

\begin{multicols}{2}
\tableofcontents[]
\end{multicols}


# Control Structures

## Background

### Control Structures and Flow of Control

- A **control structure** is a piece of code whose **analysis of a variable** 
  results in a **choice being made as to the direction** the program should go.

- Meanwhile, the **Flow of Control** specifies the **direction** the program 
  takes or **flows** when given conditions and parameters.

### Types of Control Structures:

There are **three** different types of control structures. 

- **Sequential:** Executes code statements one line after another
     - Exactly adhering to an algorithm like creating a cake from a recipe.
- **Selection:** Allows for decisions between 2 or more alternative paths. 
     - Making a choice as to whether I want a [Cold Brew](http://lp.starbucks.com/coldbrew) or [Iced Coffee](http://www.starbucks.com/menu/drinks/iced-coffee/iced-coffee) from *Starbucks*.
- **Iteration:** Enables the looping or repeating of a section code multiple times.
     - Saying the same words [over and over again](https://www.youtube.com/watch?v=NP0mQeLWCCo).

## Sequential Control

### Sequential Control

Sequential control is the default operating behavior of a computer program.
The program will read each line one after another and execute it.

```{r sequential_diagram, out.width = "1.5cm", fig.retina = NULL, fig.align='center', echo = F, cache = T}
knitr::include_graphics("support/control_diagrams/sequential/sequential_diagram.pdf")
```


## Logical Operators

### Logical Operators

To control the structure, we sometimes must make **decisions** that have
different cases based on a *boolean*, `FALSE` (0) or `TRUE` (1), condition.

A few such _vectorized_ logical operators are given as:

| Operator   | Explanation                            |
|:----------:|:---------------------------------------|
| `x == y`   | `x` equal to `y`                       |
| `x != y`   | `x` not equal to `y`                   |
| `x < y`    | `x` less than `y`                      |
| `x > y`    | `x` greater than `y`                   |
| `x <= y`   | `x` less than or equal to `y`          |
| `x >= y`   | `x` greater than or equal to `y`       |
| `x %in% y` | `x` is contained in `y`                |

### Combining Logical Operators

We can combine logical operators using:

| Operator   | Explanation                            |
|:----------:|:---------------------------------------|
| `!x`       | not `x`                                |
| `xor(x,y)` | Exclusive or, only `x` or `y` is true  |
| `x | y`    | `x` or `y`  (vectorized)               |
| `x & y`    | `x` and `y` (vectorized)               |

If we only want to compare _one_ value (e.g. not vectorized), then we can use:

| Operator   | Explanation                      |
|:----------:|:---------------------------------|
| `x || y`   | `x` or `y` (not vectorized)      |
| `x && y`   | `x` and `y` (not vectorized)     |



### Venn Diagrams of Logic: X and !X


```{r venn_diagram_x_v2, out.width = "2.5cm", fig.retina = NULL, fig.align='center', echo = F, cache = T}
knitr::include_graphics("support/logical/x/venn_diagram_x_v2.pdf")
```


```{r venn_diagram_notx_v2, out.width = "2.5cm", fig.retina = NULL, fig.align='center', echo = F, cache = T}
knitr::include_graphics("support/logical/not_x/venn_diagram_notx_v2.pdf")
```


### Venn Diagrams of Logic: AND vs. OR


```{r venn_diagram_xandy, out.width = "2.5cm", fig.retina = NULL, fig.align='center', echo = F, cache = T}
knitr::include_graphics("support/logical/xny/venn_diagram_xandy.pdf")
```


```{r venn_diagram_xory, out.width = "2.5cm", fig.retina = NULL, fig.align='center', echo = F, cache = T}
knitr::include_graphics("support/logical/xuy/venn_diagram_xory.pdf")
```

### Venn Diagrams of Logic: XOR


```{r venn_diagram_xorxy, out.width = "5cm", fig.retina = NULL, fig.align='center', echo = F, cache = T}
knitr::include_graphics("support/logical/xor/venn_diagram_xor.pdf")
```


### Examples of Logical Operators

| Explanation                            |  Example                  | Result       |
|:---------------------------------------|:--------------------------|:-------------|
| `x` equal to `y`                       | `2 == 2`                  | `TRUE `      |  
| `x` not equal to `y`                   | `1 != 3`                  | `TRUE`       |  
| `x` less than `y`                      | `10 < 8`                  | `FALSE`      |
| `x` greater than `y`                   | `10 > 8`                  | `TRUE`       |
| `x` less than or equal to `y`          | `c(6, 5) <= c(6, 1)`      | `TRUE FALSE` |
| `x` greater than or equal to `y`       | `c(6, 5) >= c(6, 1)`      | `TRUE TRUE`  |
| `y` contains `x`                       | `c(1,2) %in% c(3,1)`      | `TRUE FALSE` |
| not `x`                                | `!(1 < 2)`                | `FALSE`      |
| Exclusive or, only `x` or `y` is true  | `xor(c(T,T), c(T,F))`     | `FALSE TRUE` |
|  `x` or `y`  (vectorized)              | `c(F, F) | c(T, F)`       | `TRUE FALSE` |
|  `x` and `y`  (vectorized)             | `c(F, T) & c(F, F)`       | `FALSE FALSE`|
| `x` and `y` (not vectorized)           | `(1 < 10) && (15 > 10)`   | `TRUE`       |
|  `x` or `y` (not vectorized)           | `FALSE || TRUE`           | `TRUE`       |

### Practice with Logical Operators

Consider

```r
x = seq(0, 2, by = 0.5)
y = 2*x - 1

# Scenario 1:
x < 1

# Scenario 2:
y == 1

# Scenario 3:
!(x < 1 | y == 1)
```

What is the end result?

### Short Circuit Evaluation

- The `&&` and `||` operators both have a feature called **short-circuit evaluation**.
- Consider the `&&` (`AND`) expression (`x && y`) ...
     - If `x` is `FALSE`, then there is no reason to evaluate `y` so the evaluation stops.
     - Take for example a division check for zero:
```{r, eval = F, echo = T}
x = 0; y = 4         # Define Variables
(x != 0 &&  y/x > 0) # Evaluates only x != 0
                     # so y/x never runs.
```
- Similarly, the `||` (`OR`) expression (`x || y`) ... 
      - If `x` is `TRUE`, then whole expression is `TRUE` so there is no need to evaluate `y`.
      - Only in the case when `x` is `FALSE` does `y` get evaluated.


# Selection Control


### Selection Control

- **Selection:** Allows for decisions between 2 or more alternative paths. 
     - `if`
     - `if/else`
     - `if/elseif/else`
     - `switch`
     
```{r selection_diagram, out.width = "3.5cm", fig.retina = NULL, fig.align='center', echo = F, cache = T}
knitr::include_graphics("support/control_diagrams/selection/selection_diagram.pdf")
```

## `if` and `if/else`

     
### Selection Control - `if` and `if/else`

The `if` statement provides the ability to execute code only when the 
`condition` is `TRUE` like so:

```r
if (condition) {# True Case
  # statement
}
```

Traditionally, one would normally use `if/else` statement to address both 
`TRUE` and `FALSE` cases:

```r
if (condition) {# True Case
  # statement
} else {         # False Case
  # statement
}
```

**Note:** The `condition` clause of an `if` statement is _not_ vectorized.


### Selection Control - `if` and `if/else` Examples

```{r, echo = T}
x = 2
if (x > 0) {  # Detect if x is a positive number
  cat(x)      # Print `x` to console 
}
```

```{r, echo = T}
if (x != 42) {   # True Case
  cat("This is not the meaning of life") 
  x = 42
} else {         # False Case
  cat("This is the meaning of life!") 
}
```

### Selection Control - `if` and `if/else`: Concept Check

Identify the problematic areas of the following `if` statements in _R_:

**Example 1:**

```r
x = 9; y = 20
if (x < 10) && (y < 35)
  cat("Bingo was his name!\n")
```

**Example 2:**

```r
x = 43
if (x == 42 || 43 || 44)
  cat("That's close to my age!\n")
```

### Selection Control - `if` and `if/else`: Concept Check (Pt 2)


**Example 3:**

```r
x = seq(0, 1, by = 0.1)
y = seq(.1, 1.1, by = 0.1)
if (x < y)
  TRUE
```

### Selection Control - Vectorized `if/else` 

In the case of the last `if` statement, we have the ability to use *R*'s
vectorized `ifelse()`

```r
ifelse(condition, TRUE-CONDITION, FALSE-CONDITION)
```

Example:

```{r, echo = T, engine='R', cache = T}
x = seq(0, 1, by = 0.2)
y = seq(-1, 3, by = 0.7)
ifelse(x < y, TRUE, FALSE)
```

**Note:** Watch out for vector recycling!

### An alternative approach: Simplify the logical vector.

_R_ has two functions that allow for the simplification of _logical_ traits:

- `any()`: Only _one_ value must be `TRUE` for the `if` to be `TRUE`.
- `all()`: All values must be `TRUE` for the `if` to be `TRUE`.

### Simplifying the logical vector

Examples:

```{r}
x = 1:5
any(x < 2)
all(x > 3)

if(all(x > 0)) cat("Greater than 0") else cat("Less than 0")
```

## `if/elseif/else` 

### Selection Control - `if/elseif/else` 

Sometimes, we may wish to group together certain conditions

```{r, eval = F, echo = T, engine='R'}
if(condition) {
    # statements
} else if (condition2) {
    # statements
} else { 
    # statements
}
```

In this case, the first statement is checked, if it does not contain the correct
value the next statement is check and so until either one of the `condition`s 
is true or it arrives at the `else` condition.

### Selection Control - `if/elseif/else` 

```{r selection_diagram_switch, out.width = "4cm", fig.retina = NULL, fig.align='center', echo = F, cache = T}
knitr::include_graphics("support/control_diagrams/selection/selection_diagram_switch.pdf")
```

### Selection Control - `if/elseif/else` 

**Example:**
```{r, echo = T, cache = T}
x = 3
if (x == 42) {
  cat('Equal\n')
} else if (x > 42) {
  cat('Greater Than\n')
} else {
  cat('Less Than\n')
}
```

### Selection Control - `if/elseif/else` 

The `if/elseif/else` could be written using only `if/else`

```{r, echo = T, cache = T}
x = 3
if (x == 42) {
  cat('Equal\n')
} else {
  if (x > 42) {
    cat('Greater Than\n')
  } else {
    cat('Less Than\n')
  }
}
```

**Discussion:** Why might structuring the `if` this way be not so ideal?

## Special Selection Ops

### Selection Control - Vectorized `if/elseif/else` 

Following in the previous examples footsteps, we can write a vectorized version
to process multiple observations via:

```{r, echo = T, cache = T}
ifelse(x == 42, 'Equal', 
       ifelse(x > 42, 'Greater Than', 'Less Than'))
```

### Selection - Ternary Operator

Previously, we saw a *vectorized* version of the ternary operator given 
by `ifelse()`.

What makes a selection statement a **ternary operator** is it is meant to be
an **inline** `if/else` statement. e.g.

```{r, eval = F, echo  =T}
if(condition) TRUE  else  FALSE
```

Consider:

```{r, eval = F, echo  =T}
x = 42
a = if(x == 42) TRUE else FALSE # Verbose
b = (x == 42)                   # Concise
all.equal(a, b)
```

## `switch` 

### Selection Control - `switch` 

Within a `switch`, the condition is evaluated as the first parameter and then
compared against the values in the case labels. 

```{r, eval = F, echo = T}
switch(type,
       case_1 = statement_1,
       case_2 = statement_2,
       statement_3 # Default / Else 
)
```

Switch are often represented as `if/elseif/else` like:

\scriptsize

```{r, eval = F, echo = T}
if (type == case_1) {
  # statement 1
} else if (type == case_2) { 
  # statement 2
} else {
  # statement 3
}
```

### Selection Control - `switch` Example

There are multiple ways to switch in R. Principally, switches either respond 
to **index** or **named condition**.

**Index:**
\scriptsize

```{r, echo = T}
switch(1, 
       "First",
       "Second")
```
\normalsize

**Named condition:**

\scriptsize
```{r,  echo = T}
switch("toad", 
       prince = "First", 
       toad = "Second", 
       "Third")
```

**Note:** Faster to use a `switch` than an `if/elseif/else` in *R*!

### Selection Control - `switch`: Concept Check

Consider the following switch:

```{r, eval = F, echo = T}
switch(2, 
       prince = "First", 
       toad = "Second", 
       "Third")
```

- What will be the output? 

### Selection Control - `switch`: Concept Check

Let's make a modification to the previous switch to use an index, e.g.

```{r, eval = F, echo = T}
switch(4,                # Switched to Numeric
       prince = "First", 
       toad = "Second", 
       "Third")
```

What will be the output? 

### Selection Control - `switch`: Concept Check

- What happens when we change to use a named condition?

```{r,  eval = F, echo = T}
switch("Fourth",           # Switched to Named
       prince = "First", 
       toad = "Second", 
       "Third")
```

- How does this jive with the previous `switch` example?

# Iteration Control

### Iteration Control

- **Iteration:** Enables the looping or repeating of a section code multiple times.
     - `for`
     - `while`
     - `do/while`

### Iteration Control

> "I choose a lazy person to do a hard job. Because a lazy person will find an easy way to do it."
>
> --- Bill Gates

Iteration is the advent of the ability to repeat statements like:

```{r echo = T, eval = F}
cat("Hello World!\n")
cat("Hello World!\n")
cat("Hello World!\n")
```

Without having to type it!

## `for` loops

### Iteration Control - `for`

The most common loop in _R_ is that of the `for` loop. The `for` loop
is structured as:

```r
for(variable in vector) { 
 # statements
}
```

Typically, the `vector` component is called the _loop index_ and is an 
_integer_. However, _R_ also allows it to be vector names or actual data.

### Iteration Control - `for`: Summation Example

Consider the summation example: $\sum\limits_{i = 1}^n {{x_i}}$

```{r, echo = T, eval = F}
n = 5                     # Amount to iterate over
set.seed(111)             # Set seed for reproducibility
x = runif(n)              # Generate n data points
summed = 0                # Output Variable
for (i in 1:n) {          # Iteration Sequence
  summed = summed + x[i]  # Statement
}
```

### Iteration Control - `for`

The `for` loop provides the ability to **automatically** increment the counting
variable or iterate through a set of named expressions.

```{r, echo = T, eval = F}
x = c("coffee","doppio espresso", 
      "iced coffee", "cold brew")

# Numeric Index
for (i in 1:length(x)) {
  cat("i is: ", i, ", x[i] is:", x[i],"\n", sep= "")
}

# Named Index
for (i in x) {
  cat("i is: ", i, "\n", sep= "")
}
```


### Iteration Control - Index Examples

Consider the loop of:

```r
a = numeric(0)
for(i in 1:length(a)){
  cat("Hello!\n")
}
```

1. What _should_ the output be? 
2. What _is_ the output? 


### Iteration Control - Index Protection

Instead of using `1:length(a)` to iterate a loop counter, define `n = length(a)`
and try to use: `seq_len(n)` or `seq_along(a)`

```r
a = numeric()
n = length(a)

# Case 1
for(i in 1:n) { cat("Hello Unprotected Loop") }

# Case 2
for(i in seq_len(n)) { cat("Hello Protected Loop v1!\n") }

# Case 3
for(i in seq_along(a)) { cat("Hello Protected Loop v2!\n") }
```

### Iteration Control - Index Protection

- In essence, the error in **Case 1** is due to `length(a)` being `0`. 
- This leads the sequence `1:length(a)` to be `1:0`, which creates the
  vector `c(1,0)`. 
- In this case, the `for` loop is then able to iterate.
    - If we had decided to _subset_ a value, then an error would be thrown!

### Iteration Control - `for` skipping

Sometimes, we might have a case that we want to *skip* in the loop. To do so,
we use the `next` statement to move the loop forward.

```{r cache = T}
for (i in 1:4) {
  if(i %% 2 == 1) {
    next      # skip odd numbers
  }
  cat(i, "\n")
}
```

## `while` loops

### Iteration Control - `while`

Meanwhile, a `while` loop requires three items:

1. condition, 
2. statements,
3. change logic

```{r, echo = T, eval = F}
while (condition) {
  # statements
  
  # change logic
}
```

### Iteration Control - `while`

Take for example the following loop:

```{r, echo = T, eval = F}
i = 1
while (i < 3) {
  cat(i, "\n")
  i = i + 1    # Do not forget to increment!
}
```

This is a perfect candidate for looping as long as we remember to **increment**
the looping variable.

### Checking in with a `while`

What is the output of the following piece of code?

\scriptsize

```{r example_while, echo = T, eval = F}
y = 0
x = 10
i = 0

while(i < x) {
  if(i %% 2 == 0) {
    i = i + 1
  } else if (i %%3 == 0) {
    i = i + 2
  } else {
    y = y + i
  }
}

y == 18
```

### Iteration Control - `break` out from a loop

We may obtain the desired result before the end of the loop and, thus, no
longer need to continue the iteration.

To escape from a loop, we need to use `break`.^[**Note:** Break also works in a `for` loop.]

```{r, echo = T, eval = F}
i = 1
while (i < 5) {
  if(i == 3) { # Stops the loop at the 3rd iteration.
    break
  }
  cat(i, "\n")
  i = i + 1    # Do not forget to increment!
}
```

## `repeat` loops

### Iteration Control - `repeat` or a `do-while`

Previously, for both `for` and `while` we had to have the initial condition met
before the loop would run. However, sometimes we may want to let code run
until a specific threshold is met (e.g. for a simulation). 

\scriptsize
```{r, echo = T, eval = F}
x = 1
y = 10
ntrials = 0
repeat {
  
  if(runif(1) > 0.5) {
    x = x + 1  
  }
  
  ntrials = ntrials + 1
  
  if(x < y) {
    break
  }
}
```

What is the output of `ntrials < x`?


### Summary of Control Structures

- Discussed logical conditions
- Explored the different levels of selection controls `if`, `if/else`, `if/elseif/else`, `switch`
- Talked about the different types of R loops: `for`, `while`, and `repeat`
- Emphasized proper loop indexing techniques.
